---
title: Simple Moving Average algorithm & backtest
author: ~
date: '2019-10-12'
slug: simple-moving-average-algorithmic-backtest
categories: ['Analysis']
tags: ['R', 'algotrading', 'ML', 'prediction']
disableComments: no
bestof: yes
summary: 'I made an algorithmic trader in R based on the "moving average crossover" technical indicator and performed a backtest using the TSLA stock. It did not perform well, but it was an interesting challenge that involved some timeseries analysis.'
---

I decided to try making a simple algorithm trader in R.

The algorithm is based on the "moving average crossover". Basically, the idea is to take two simple moving averages (SMAs) -- one that averages over a longer time window, and one that averages over a shorter time window -- and buy or sell based on when the two SMAs crossover. Here I used 50-day and 200-day SMAs. Buy decisions were when the SMA-50 went above the SMA-200, and sell decisions were when the SMA-50 went below the SMA-200.

To obtain historical stock data, I used the <a href="https://github.com/business-science/alphavantager">alphavantager</a> R package to access the <a href="https://www.alphavantage.co/documentation/">Alpha Vantage API</a>; the <a href="https://github.com/business-science/tidyquant">tidyquant</a> package to calculate simple moving averages; and tidyverse, ggplot2, etc to wrangle and visualize data.

I picked Tesla (ticker: "TSLA") as the stock I would invest in, because as I understand it has had a lot of volatility which makes it fruitful for day trading. I assumed a $100 USD investment and simulated a backtest over the daily close data between 2015 and 2019. <strong>TL;DR: My algorithm did not make money.</strong> But this was a fun exercise nonetheless!

***

# Setup

Load libraries.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidyquant)
library(reshape2)
library(ggplot2)
library(lubridate)
```

```
library(tidyverse)
library(tidyquant)
library(alphavantager)
library(reshape2)
library(ggplot2)
library(lubridate)
```

Set Alphavantager key.

```{r}
# av_api_key(readLines("C:/Users/tyler/Documents/av_key.txt"))
```

# Get historical stock data

Get daily price data between 2016 and 2019.

```{r}
# av_get(symbol = "TSLA",
#        av_fun = "TIME_SERIES_DAILY",
#        outputsize = 'full') %>%
#   select(timestamp, close, high, low) %>%
#   filter(timestamp > as.Date('2015-01-01') &
#          timestamp < as.Date('2019-12-31')) -> df
# save(df, file = "tsla_price_data.RData")

load("tsla_price_data.RData")
```

# Simple Moving Averages

Calculate simple moving averages (50-day and 200-day).

```{r}
df %>%
  mutate(sma_50 = SMA(close, n = 50),
         sma_200 = SMA(close, n = 200)) %>%
  filter(!is.na(sma_200)) -> df_with_sma
```

Convert long to wide.

```{r}
melt(df_with_sma, id.vars = c("timestamp")) -> df_long
```

Plot daily close, SMA50 and SMA200.

```{r}
df_long %>%
  filter(variable %in% c('close', 'sma_50', 'sma_200')) %>%
  ggplot(., aes(x = timestamp, y = value)) + 
    geom_line(aes(color = variable), size = 1) + 
    ggtitle("TSLA stock with SMAs")
```

# Encode buy/sell signals and decisions

We'll use SMA50/SMA200 crossover events to make buy/sell decisions.

- Buy if SMA50 > SMA200 (unless we've bought already)
- Sell if SMA50 < SMA200 (unless we've sold already)

We'll encode the decisions in two passes: the first pass will encode the signal, and the second pass will encode the decision. Basically, the decision encoding will make sure we won't buy again if the last action was to buy, and we won't sell again if the last action was to sell.

```{r}
df_with_sma %>%
  mutate(
    # Buy if SMA50 > SMA200
    # Otherwise sell
    signal = case_when(
      sma_50 > sma_200 ~ "buy",
      sma_50 < sma_200 ~ "sell"
    ),
    # Create a lagging indicator for action
    previous_signal = lag(signal, 1),
    # Encode hold as any time when current and last action are the same
    # since we don't want to keep buying / selling
    decision = case_when(
      signal == previous_signal ~ "hold",
      TRUE ~ signal
    )
  ) -> df_with_decisions
```

Extract the periods when we would have bought or sold.

```{r}
df_with_decisions %>%
  filter(decision != 'hold') %>%
  select(timestamp, close, high, low, sma_50, sma_200, decision) -> df_bought_sold

df_bought_sold
```


# Visualizing the algorithm decisions

Visualize the decisions by plotting them as vertical intercept lines over the graph.

```{r}
df_long %>%
  filter(variable %in% c('close', 'sma_50', 'sma_200')) %>%
  ggplot(., aes(x = timestamp, y = value)) + 
  geom_line(aes(color = variable), size = 1) +
  geom_vline(data = df_bought_sold %>% filter(decision == 'buy'), 
             aes(xintercept = as.numeric(timestamp)),
             col = 'black',
             linetype="dashed") +
  geom_vline(data = df_bought_sold %>% filter(decision == 'sell'), 
             aes(xintercept = as.numeric(timestamp)),
             col = 'red') + 
  ggtitle("TSLA stock with SMAs and algorithmic buy/sell decisions\n(black = buy, red = sell)")
```

# Simulating the backtest

Calculate gains/losses, assuming 100 initial investment, and assuming we buy/sell at the mid-day price.

```{r}
usd <- 100
shares <- 0
stock_price <- NA
for(i in 1:nrow(df_bought_sold)){
  stock_price <- (df_bought_sold[i,]$high + df_bought_sold[i,]$low)/2
  
  if(shares > 0 && df_bought_sold[i,]$decision == "sell") {
    
    print(paste0("SELL: ", shares, " shares @ $", stock_price, " USD"))
    usd <- stock_price * shares
    print(paste0("USD IS NOW @ ", round(usd, 2)))
    shares <- 0
    
  } else if(usd > 0 && df_bought_sold[i,]$decision == "buy"){
    
    print(paste0("BUY: ", shares, " shares @ $", stock_price, " USD"))
    stock_price <- df_bought_sold[i,]$close
    shares <- usd/stock_price
    usd <- 0
    
  }
  
  cat("\n")
  
}
```

# Final result

What was our total ROI?

```{r}
total_return <- usd/100
print(paste0("Total return: ", round(total_return*100)-100, "%"))
```

Wow that did not turn out well!




