---
title: Implicit datatypes and coercion are weird
author: ~
date: '2019-10-05'
slug: implicit-datatypes-and-coercion-are-weird
categories: ['Other']
tags: ['R', 'pyMTurkR']
disableComments: no
---

The other day, a user of my <a href="https://github.com/cloudyr/pyMTurkR">pyMTurkR</a> package wrote to say they encountered an issue with the `GetAssignments()` function. When they tried calling the function it would throw the following error:

```
Error in as.POSIXlt.character(x, tz, ...) :
  character string is not in a standard unambiguous format
```

In the process of debugging this error, I learned a lesson about how R dataframes behave: Unless a dataframe's columns are explicitly defined to have a datatype, they can change to accommodate the type of data that is first put into them.

## R dataframes are flexible

The dataframe is an R object with columns and rows, where each column of a dataframe has a single data type. For example, a dataframe might have a column `name` that contains data of the `character` type, or a column `age` that contains data of the `double` type (a type of number). But dataframes are flexible, and the data type doesn't have to be defined when a dataframe is created. 

For example, you can create a dataframe with no rows or columns. Then you can put stuff into the dataframe by binding to it the rows of another dataframe that contains data. If a dataframe is empty, you can bind the rows of any other dataframe to it.

```{r}
df <- data.frame()
df

df <- rbind(df, data.frame(greeting = 'Hello!'))
df
```

You can also create a dataframe that has columns but no rows. This can be done by giving the dataframe a matrix with predefined dimensions, like this

```{r}
data.frame(matrix(NA, nrow = 0, ncol = 2), stringsAsFactors = FALSE)
```

You can also name the columns of the empty dataframe using the `setNames` function from the `stats` library, like this:

```{r}
colnames <- c('name', 'age')
stats::setNames(data.frame(matrix(NA, nrow = 0, ncol = 2), stringsAsFactors = FALSE), colnames)
```

This can be helpful if you want to initialize a dataframe with pre-defined columns to put data into later. If you check the column types for the dataframe just created, you'll find that they were initialized to `logical` because the matrix contained `NA` which is a logical type of data.

```{r}
typeof(NA)
df <- setNames(data.frame(matrix(NA, nrow = 0, ncol = 2), stringsAsFactors = FALSE), colnames)
typeof(df$name)
```

If you define the names of the dataframe when it's initialized, this will have the effect of constraining the data that can be put into it. For example, if you define a dataframe in advance that has 2 columns, then you can iterate over the dataframe and put data into each column by referring to the indices `1` and `2`.

```{r}
colnames <- c('weekday', 'hour_of_day')
df <- stats::setNames(data.frame(matrix(NA, nrow = 0, ncol = 2), stringsAsFactors = FALSE), colnames)
for(i in 1:3){
  df[i,1] <- sample(weekdays(.leap.seconds), 1) # weekday
  df[i,2] <- round(runif(min = 0, max = 24, n = 1)) # hour_of_day
}
df
```

Importantly, the dataframe is flexible. Even though the columns were initialized to have a `logical` type (because the matrix contained NA), when the dataframe is initialized it doesn't yet contain any data, so we're not confined to putting only `logical` data into those columns.

If we put a different type of data into one of these columns, R knows that it should probably change the column's data type to match whatever we put into it. For example, if we `rbind` the empty dataframe to one containing a name of `character` type and an age of `double` type, the name and age columns will change to assume these types accordingly.

```{r}
colnames <- c('name', 'age')
df <- setNames(data.frame(matrix(NA, nrow = 0, ncol = 2), stringsAsFactors = FALSE), colnames)
typeof(df$name)
typeof(df$age)

df <- rbind(df, data.frame(name = 'Tyler', age = 33))
typeof(df$name)
typeof(df$age)
```

This flexibility is great and usually helpful. But it also creates the potential for unexpected behavior.

## Implicit dataframe datatypes can sometimes get weird

### Example 1

Imagine you have an API for an online shopping sites that returns the following information about orders:

- Name on order
- Time order was placed
- Time order was fulfilled

In the form of a dataframe, a single item might look like this:

```{r}
data.frame(
  name = "Tyler",
  time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(2),
  time_filled = as.POSIXct(Sys.Date()) + lubridate::hours(2)
)
```

If you wanted to ingest a lot of orders and then put them into the same dataframe, you might start by initializing an empty dataframe like we did before using the matrix method.

```{r}
colnames <- c('name', 'time_placed', 'time_filled')
df <- stats::setNames(data.frame(matrix(NA, nrow = 0, ncol = 3), stringsAsFactors = FALSE), colnames)
df
```

Then, you can rbind in the data for each order. (Imagine the `orders` object here was given to us by an API.)

```{r}
order1 <- data.frame(
            name = "Tyler",
            time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(2),
            time_filled = as.POSIXct(Sys.Date()) + lubridate::hours(2)
          )

order2 <- data.frame(
        name = "Friedrich",
        time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(6),
        time_filled = as.POSIXct(Sys.Date()) + lubridate::hours(1)
      )

orders <- list(order1, order2)

df_final <- df
for(order in orders){
  df_final <- rbind(df_final, order)
}
df_final

typeof(df_final$time_filled)
```

But what happens if each of the orders doesn't contain the same type of data? For example, what if the fulfillment time (`time_fulfilled`) is `NA` when the order has not yet been filled? If the first order that we put into the dataframe has the data type we're expecting -- a `POSIXct` time/date type) -- then the column gets initialized to a `POSIXct` column and there's no problem if `NA` values come in later because `POSIXct` data can assume an `NA` value.

```{r}
df_final <- df # df is our empty dataframe from before

order1 <- data.frame(
            name = "Tyler",
            time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(2),
            time_filled = as.POSIXct(Sys.Date()) + lubridate::hours(2)
          )

order2 <- data.frame(
        name = "Friedrich",
        time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(6),
        time_filled = as.POSIXct(Sys.Date()) + lubridate::hours(1)
      )

orders <- list(order1, order2)

for(order in orders){
  df_final <- rbind(df_final, order)
}
df_final

typeof(df_final$time_filled)
```

If, on the other hand, the first order that we put into the dataframe contains an `NA` value for that column, the column first gets initialized to a `logical`. We saw this in the previous section, no big deal.

```{r}
df_final <- df # df is our empty dataframe from before

order1 <- data.frame(
            name = "Tyler",
            time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(2),
            time_filled = NA
          )

df_final <- rbind(df_final, order1)
df_final

typeof(df_final$time_filled)
```

But -- and here's where it gets a bit weird -- if the next order that comes in has a value for that same column, the column gets coerced to a `double` and we see a numerical value instead (this is what we call an `epoch` value). For some reason, R has decided that this is the best data type to use for the column. 

```{r}
order2 <- data.frame(
        name = "Friedrich",
        time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(6),
        time_filled = as.POSIXct(Sys.Date()) + lubridate::hours(1)
      )

df_final <- rbind(df_final, order2)
df_final
```

Instead of initializing the matrix with `NA`, we also could have used an empty string (`""`) for the same result.

```{r}
colnames <- c('name', 'time_placed', 'time_filled')
df <- stats::setNames(data.frame(matrix("", nrow = 0, ncol = 3), stringsAsFactors = FALSE), colnames)
df

df_final <- df # df is our empty dataframe from before

order1 <- data.frame(
            name = "Tyler",
            time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(2),
            time_filled = NA
          )

df_final <- rbind(df_final, order1)
df_final

order2 <- data.frame(
        name = "Friedrich",
        time_placed = as.POSIXct(Sys.Date()) - lubridate::hours(6),
        time_filled = as.POSIXct(Sys.Date()) + lubridate::hours(1)
      )

df_final <- rbind(df_final, order2)
df_final
```

### Example 2

But here's where it gets a bit weird. If we initialize the dataframe with an empty string (`""`) and then add data by row and column indices, but only for the elements that don't contain an `NA` value, then we get a different datatype altogether.

```{r}
colnames <- c('name', 'time_placed', 'time_filled')
df <- stats::setNames(data.frame(matrix("", nrow = 0, ncol = 3), stringsAsFactors = FALSE), colnames)
df

df_final <- df # df is our empty dataframe from before

df_final[1,1] <- "Tyler"
df_final[1,2] <- as.POSIXct(Sys.Date()) - lubridate::hours(2)
# df_final[1,3] Because time_filled was empty, we'll skip adding it to the dataframe and let it assume a value

df_final

typeof(df_final$time_filled)

df_final[2,1] <- "Tyler"
df_final[2,2] <- as.POSIXct(Sys.Date()) - lubridate::hours(6)
df_final[2,3] <- as.POSIXct(Sys.Date()) - lubridate::hours(1)

df_final
```

This time `time_filled` becomes a `character` type!

## POSIXct coercion is also weird

Coercing `POSIXct` is also a bit weird. `POSIXct` data is stored in epoch time, which is the number of seconds since the year 1970. It can be converted to `numeric` or `character` types, and then converted back. Although when converting a number back the "origin" has to be supplied. 

The `numeric` form can be also converted to a `charcter` string, but it can't be convered back to `POSIXct` at that point without first converting it back to `numeric`.

```{r}
time <- as.POSIXct(Sys.Date())

as.character(time) -> chr
chr
as.POSIXct(chr)

as.numeric(time) -> num
num
as.POSIXct(num, origin = '1970-01-01')

as.character(num) -> numchr
numchr
try(as.POSIXct(numchr))
```

```{r}
as.POSIXct(as.numeric(numchr), origin = '1970-01-01')
```


## Insight into the original bug report

As I was investigating the problem underlying the bug that would produce the error, I realized that implicit datatyping and POSIXct were both to blame.

```
Error in as.POSIXlt.character(x, tz, ...) :
  character string is not in a standard unambiguous format
```

The `time` column was sometimes `POSIXct` and other times it was an empty `character` string. 

If you try joining these together, you get an error if the first value that comes in is of the `POSIXct` type, because a `character` cannot be coerced to a `POSIXct`, but the reverse isn't possible.

```{r}
data.frame(name = "Tyler", time = "", stringsAsFactors = F) -> a
data.frame(name = "Friedrich", time = as.POSIXct(Sys.Date()), stringsAsFactors = F) -> b
rbind(a, b)
try(rbind(b, a))
```

The solution was to catch the empty values as they were coming in and convert them to `NA` of the `POSIXct` type.

```{r}
time <- ""
if(time == "") {
  data.frame(name = "Tyler", time = as.POSIXct(NA)) -> a
}

time <- as.POSIXct(Sys.Date())
data.frame(name = "Friedrich", time = time) -> b

rbind(a, b)
rbind(b, a)
```

