---
title: "Macronutrient Analysis of Recipes Using Natural Language Processing API"
author: "Tyler Burleigh"
date: "3/9/2019"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Knowing is half the battle

Have you ever wanted to know the macronutrient composition of a recipe? How many
carbs are you *really* consuming in that sandwich? How much protein are you *really*
getting from a plate of beans?

Now you can find out!

## What my Shiny app does

I created a Shiny app that sends a user input to the Nutritionix API.
https://www.nutritionix.com

This API uses natural language processing to identify the ingredients, and the
quantity of ingredients in the recipe. It returns a list with nutritional data
about each ingredient. My app then calculates a sum total for

- Fat
- Protein
- Carbohydrates
- Calories

And returns this to the user.

## The app

<iframe src="https://wizzywig.shinyapps.io/nutrition/" height="100%" width="75%" style="border:0"></iframe>

## Recommended daily intake

For a diet with 1,800 calories per day, it is recommended that you consume the following:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(plotly)
p <- plot_ly(
  x = c("fat", "protein", "carbs"),
  y = c(65, 145, 290),
  text = c(65, 145, 290), 
  textposition = 'auto',
  name = "Recommended Macro Intake (grams)",
  type = "bar"
)
p
```

## Thank you

I hope you enjoyed this presentation. Thank you, and stay healthy!

(Please note: This slideshow is hosted on GitHub pages, but I am using a custom domain name. You can find the source here:
https://github.com/tylerburleigh/tylerburleigh.github.io/tree/master/DDP/pitch_slideshow)



