---
date: ""
disableComments: true
slug: ""
title: About Me
output: html_document
---

For more than ten years, I've been working at the intersection of behavior change and data science. From academia to healthcare tech startups to being a part of the industry's first Behavioral Science team, I've been applying rigorous scientific methods and data analytics to improve decision-making in business, always with the goal of driving towards decisions that make the world a better place.


You can find the best examples of my work here on <a href="https://tylerburleigh.com">my Blog</a>, on my <a href="https://tylerburleigh.com/page/writings">Publications page</a>, and at <a href="https://github.com/tylerburleigh">my GitHub</a>.

# Visual timeline

```{r, echo=FALSE, message=F, warning=F}
library(timevis)

data <- data.frame(
  id      = 1:8,
  
  content = c("BA in Psychology", 
              "MA in Cognitive Psychology",
              "PhD in Applied Cognitive Science",
              "Assistant Research Scientist @ New York University",
              "Research Scientist @ Datacubed Health",
              "Data Scientist @ Clover Health",
              "Data Science Specialization @ Coursera",
              "Data Scientist with Python @ Datacamp"),
  
  start   = c("2005-09-01", 
              "2009-09-01",
              "2011-09-01",
              "2017-02-28",
              "2017-12-31",
              "2018-11-15",
              "2018-07-01",
              "2019-12-01"),
  
  group = c(1, 1, 1, 2, 2, 2, 1, 1)
)

timevis(data, groups = data.frame(id = 1:2, content = c("G1", "G2"))) %>%
  setGroups(data.frame(id = 1:2, content = c("Education", "Jobs"))) %>%
  setOptions(options = list(showCurrentTime = FALSE, orientation = "top"))
```

Code for the above visualization:

```{r, eval=F}
library(timevis)

data <- data.frame(
  id      = 1:8,
  
  content = c("BA in Psychology", 
              "MA in Cognitive Psychology",
              "PhD in Applied Cognitive Science",
              "Assistant Research Scientist @ New York University",
              "Research Scientist @ Datacubed Health",
              "Data Scientist @ Clover Health",
              "Data Science Specialization @ Coursera",
              "Data Scientist with Python @ Datacamp"),
  
  start   = c("2005-09-01", 
              "2009-09-01",
              "2011-09-01",
              "2017-02-28",
              "2017-12-31",
              "2018-11-15",
              "2018-07-01",
              "2019-12-01"),
  
  group = c(1, 1, 1, 2, 2, 2, 1, 1)
)

timevis(data, groups = data.frame(id = 1:2, content = c("G1", "G2"))) %>%
  setGroups(data.frame(id = 1:2, content = c("Education", "Positions"))) %>%
  setOptions(options = list(showCurrentTime = FALSE, orientation = "top"))
```

